An Efficient Calculus for Generalised Parikh Images of Regular Languages

### 2022-07-05 main.tex
Update data, extensions

\begin{definition}
  A \textit{cut}, $C$ of an automaton~$\Automaton = \AutomatonTuple$,
  $\SeparatingCut(C, \Automaton, \State)$, is a minimal set of transitions whose
  removal from~$\Transitions$ would cause~$\State \in \States$ to be unreachable
  from the initial state~$\InitialState$. $\SeparatingCut(\Automaton,
  \InitialState)$ is crucially undefined since it cannot be made unreachable
  from itself by definition.
\end{definition}

@@

@@

  We define $\Filter$ after substitutions from both initial applications of
  $\FlowEq$ as follows:

% Propagated equations
    \begin{equation*}
      \begin{aligned}
        & \Filter(\FromLabelTo{start}{a}{sawA}) & = 1 - \TransitionVar_6 \\
        & \Filter(\FromLabelTo{start}{b}{sawB}) & = \TransitionVar_4 + \TransitionVar_6 \\
        & \Filter(\FromLabelTo{sawA}{c}{sawA})  & = \TransitionVar_3  \\
        & \Filter(\FromLabelTo{sawB}{b}{start}) & = 2\TransitionVar_6 + \TransitionVar_4 \\
        & \Filter(\FromLabelTo{sawA}{a}{final}) & = 1 - \TransitionVar_6 \\
        & \Filter(\FromLabelTo{sawB}{c}{final}) & = \TransitionVar_6 \\
        % Other automaton
        & \Filter(\FromLabelTo{start}{\Sigma}{start}) & = \TransitionVar_7 \\
        & \Filter(\FromLabelTo{start}{c}{final}) & = 1 \\
        & \Filter(\FromLabelTo{final}{\Sigma}{final}) & = \TransitionVar_9 \\
      \end{aligned}
    \end{equation*}

\begin{prooftree}
  \infer0[]{
    \begin{matrix}
      \begin{bmatrix}
        a \\
        b \\
        c
        \end{bmatrix} = \begin{bmatrix}
          2 - 2\TransitionVar_6 \\
          \TransitionVar_4 + \TransitionVar_6 + 2\TransitionVar_6 + \TransitionVar_4 \\
          \TransitionVar_3 + \TransitionVar_6
        \end{bmatrix}
        \\
        \begin{bmatrix} % What to do about sigma!?
          a \\
          b \\
          c
          \end{bmatrix} = \\
      \Connected(\AcaOrBc{}, \Filter) \land \\
      \Connected(\SomethingCSomething{}, \Filter) \land \\
      \Image{}_{\AcaOrBc{}\times\SomethingCSomething{}, \Map}(\Filter, 
      \begin{bmatrix}
        a \\
        b \\
        c
        \end{bmatrix}) \land \\
        a > b
    \end{matrix}
  }
  \infer1[\ExpandM]{\Image{}_{\AcaOrBc{}\times\SomethingCSomething{}, \Map}(\Filter, \begin{bmatrix}
    a \\
    b \\
    c
    \end{bmatrix}) \land a > b}
\end{prooftree}

@@

The other source of clauses to learn is the \Materialise{} rule. Whenever an
attempt to materialise a product of two ($\Filter$-filtered) automata
$\Automaton_1 \times \Automaton_2$ produces an empty product, we can determine
the cause of the failure with respect to the automata and their respective
transition variables. \Fudge{How? I have no idea}

\contents{\item dela upp på flera predikat
\item lär no-good kombinationer}

@@

Extending \Calculus{} to support fully symbolic \Fudge{automata is straigthforward}. The
only difficulty consists in handling the mapping of the homomorphism $\Map$ over
symbolic labels, assuming it maps to finitely many monoid elements.

\Fudge{We should probably expand here? I don't know}

@@

@@

The benchmarks are executed in parallel, since they are mostly single-threaded and initial results showed negligible interference on performance, on \BenchmarkRig{}. We compiled the code using Scala~\ScalaVersion{}, and executed the experiments on~\JvmVersion{} with a maximum heap of~\MaxHeapSize{}. We used \Nuxmv{} version~\NuxmvVersion{} invoked as a subprocess for each instance. Experiments were executed on a fresh JVM per backend, but the JVM was kept active for all benchmarks, meaning that JIT compilation would make subsequent executions faster. We believe this represents a realistic use case where \Calculus{} is integrated into a string solver or other system and be called repeatedly. To mitigate systemic impacts of this decision on our results, experiments were executed in random order for all backends. We run all  with a timeout of~\RuntimeTimeout{}.

As a way of gauging the differences in overhead between the native backends and \Nuxmv, we executed a small experiment where we ran the same trivial instance 30 times and observed the runtimes for all solvers. All backends saw an improvement from JIT compilation within the first three runs, but subsequent runs only improved the native backends. The final measured overhead for the fully warm JVMs was less than~\OverheadSeconds. From this we draw the conclusion that major differences in runtime between the native backends and \Nuxmv{} is more likely to stem from innate differences rather than unfair overhead resulting from our particular configuration.

@@

In Figures~\ref{fig:solve-division} we show how many of the~\NrBenchmarks{} the respective back-ends could solve and with which status. A full summary of their outcomes is also available in Table~\ref{tab:solve-status}. We see here \Fudge{that some solver is better than the others}.

>>>>

\begin{definition}
  A \textit{cut}, $C$ of an automaton~$\Automaton = \AutomatonTuple$,
  $\SeparatingCut(C, \Automaton, \State)$, is a minimal set of transitions whose
  removal from~$\Transitions$ would cause~$\State \in \States$ to be unreachable
  from the initial state~$\InitialState$, or cut it off from reaching any
  accepting state. $\SeparatingCut(\Automaton, \InitialState)$ is crucially
  undefined if $\Automaton$ accepts any string since it cannot be made unreachable
  from itself by definition and must have at least onepath to an accepting state
  if it accepts any string.
\end{definition}

@@

\subsection{An Example}\label{sec:multiple:example}

@@

  We define $\Filter$ after substitutions from both initial applications of
  $\FlowEq$ as follows, note the element-wise expansion for the $\Sigma$ labels:

% Propagated equations
    \begin{equation*}
      \begin{aligned}
        & \Filter(\FromLabelTo{start}{a}{sawA}) & = 1 - \TransitionVar_6 \\
        & \Filter(\FromLabelTo{start}{b}{sawB}) & = \TransitionVar_4 + \TransitionVar_6 \\
        & \Filter(\FromLabelTo{sawA}{c}{sawA})  & = \TransitionVar_3  \\
        & \Filter(\FromLabelTo{sawB}{b}{start}) & = 2\TransitionVar_6 + \TransitionVar_4 \\
        & \Filter(\FromLabelTo{sawA}{a}{final}) & = 1 - \TransitionVar_6 \\
        & \Filter(\FromLabelTo{sawB}{c}{final}) & = \TransitionVar_6 \\
        % Other automaton
        & \Filter(\FromLabelTo{start}{\Sigma}{start}) & = \TransitionVar_{7a}, \TransitionVar_{7b}, \TransitionVar_{7c} \\
        & \Filter(\FromLabelTo{start}{c}{final}) & = 1 \\
        & \Filter(\FromLabelTo{final}{\Sigma}{final}) & = \TransitionVar_{9a}, \TransitionVar_{9b}, \TransitionVar_{9c} \\
      \end{aligned}
    \end{equation*}

\begin{prooftree}
  \infer0[\PresburgerClose{}]{
    \begin{matrix}
      1 = \TransitionVar_{10} \\
      \TransitionVar_3 = \TransitionVar_{11} \\
      1 = \TransitionVar_{12} \\
      1 = \TransitionVar_{11} \\
      2 > 0
    \end{matrix}  
  }
  \infer1[\Subsume{}, \Expand{}, \Subsume{}]{
    \begin{matrix}
      1 = \TransitionVar_{10} \\
      \TransitionVar_3 = \TransitionVar_{11} \\
      1 = \TransitionVar_{12} \\
      1 = \TransitionVar_{11} \\
      \Image{}_{\Automaton', \Map}(\Filter, 
      \begin{bmatrix}
        2 \\
        0 \\
        1
        \end{bmatrix}) \land \\
      \Connected(\SomethingCSomething{}, \Filter) \land \\
        2 > 0
    \end{matrix}  
  }
  \infer1[Eq. reasoning]{
    \begin{matrix}
      1 = \TransitionVar_{10} \\
      \TransitionVar_3 = \TransitionVar_{11} \\
      1 = \TransitionVar_{12} \\
      1 = \TransitionVar_{11} \\
      \Image{}_{\Automaton', \Map}(\Filter, 
      \begin{bmatrix}
        a \\
        0 \\
        c
        \end{bmatrix}) \land \\
      \TransitionVar_6 = 0 \land \\
      \TransitionVar_4 = 0 \land \\
      2\TransitionVar_6 + \TransitionVar_4 = 0 \land \\
      \begin{bmatrix}
        a \\
        b \\
        c
        \end{bmatrix} = \begin{bmatrix}
          2  \\
          0 \\
          1
        \end{bmatrix} \land
        \\
        \begin{bmatrix}
          a \\
          b \\
          c
          \end{bmatrix} = \begin{bmatrix}
            \TransitionVar_{7a} + \TransitionVar_{9a} \\
            0 \\
            \TransitionVar_{7c} + \TransitionVar_{9c} + 1
          \end{bmatrix} \land \\
      \Connected(\SomethingCSomething{}, \Filter) \land \\
        a > 0
    \end{matrix}  
  }
\end{prooftree}

\begin{prooftree}
  \infer0[\Materialise]{
    \begin{matrix}
      \TransitionVar_6 = 0 \\
      \TransitionVar_4 = 0 \\
      2\TransitionVar_6 + \TransitionVar_4 = 0 \\
      \begin{bmatrix}
        a \\
        b \\
        c
        \end{bmatrix} = \begin{bmatrix}
          2  \\
          0 \\
          \TransitionVar_3 
        \end{bmatrix}
        \\
        \begin{bmatrix}
          a \\
          b \\
          c
          \end{bmatrix} = \begin{bmatrix}
            \TransitionVar_{7a} + \TransitionVar_{9a} \\
            0 \\
            \TransitionVar_{7c} + \TransitionVar_{9c} + 1
          \end{bmatrix} \\
      \Connected(\SomethingCSomething{}, \Filter) \land \\
      \Image{}_{\AcaOrBc{}\times\SomethingCSomething{}, \Map}(\Filter, 
      \begin{bmatrix}
        a \\
        0 \\
        c
        \end{bmatrix}) \land \\
        a > 0
    \end{matrix}  
  }
  \infer1[\Subsume{}]{
  \begin{matrix}
    \TransitionVar_6 + \TransitionVar_4 = 0 \\
    1 - \TransitionVar_6 > 0 \\
    \TransitionVar_6 = 0 \\
    \TransitionVar_4 = 0 \\
    \TransitionVar_3 > 0 \\
    2\TransitionVar_6 + \TransitionVar_4 = 0 \\
    \begin{bmatrix}
      a \\
      b \\
      c
      \end{bmatrix} = \begin{bmatrix}
        2  \\
        0 \\
        \TransitionVar_3 
      \end{bmatrix}
      \\
      \begin{bmatrix} % What to do about sigma!?
        a \\
        b \\
        c
        \end{bmatrix} = \begin{bmatrix}
          \TransitionVar_{7a} + \TransitionVar_{9a} \\
          0 \\
          \TransitionVar_{7c} + \TransitionVar_{9c} + 1
        \end{bmatrix} \\
    \Connected(\AcaOrBc{}, \Filter) \land \\
    \Connected(\SomethingCSomething{}, \Filter) \land \\
    \Image{}_{\AcaOrBc{}\times\SomethingCSomething{}, \Map}(\Filter, 
    \begin{bmatrix}
      a \\
      0 \\
      c
      \end{bmatrix}) \land \\
      a > 0
  \end{matrix}
  }
  \infer1[Eq. reasoning]{
    \begin{matrix}
      \begin{bmatrix}
        a \\
        b \\
        c
        \end{bmatrix} = \begin{bmatrix}
          2 - 2\TransitionVar_6 \\
          2\TransitionVar_4 + 3\TransitionVar_6 \\
          \TransitionVar_3 + \TransitionVar_6
        \end{bmatrix}
        \\
        \begin{bmatrix} % What to do about sigma!?
          a \\
          b \\
          c
          \end{bmatrix} = \begin{bmatrix}
            \TransitionVar_{7a} + \TransitionVar_{9a} \\
            \TransitionVar_{7b} + \TransitionVar_{9b} \\
            \TransitionVar_{7c} + \TransitionVar_{9c} + 1
          \end{bmatrix} \\
      \Connected(\AcaOrBc{}, \Filter) \land \\
      \Connected(\SomethingCSomething{}, \Filter) \land \\
      \Image{}_{\AcaOrBc{}\times\SomethingCSomething{}, \Map}(\Filter, 
      \begin{bmatrix}
        a \\
        b \\
        c
        \end{bmatrix}) \land \\
        a > b
    \end{matrix}
  }
  \infer1[\ExpandM]{\Image{}_{\AcaOrBc{}\times\SomethingCSomething{}, \Map}(\Filter, \begin{bmatrix}
    a \\
    b \\
    c
    \end{bmatrix}) \land a > b}
\end{prooftree}

@@

The other source of clauses to learn is the \Materialise{} rule. Whenever an
attempt to materialise a product of two ($\Filter$-filtered) automata
$\Automaton_1 \times \Automaton_2$ produces an empty product, we can determine
the cause of the failure with respect to the automata and their respective
transition variables, and learn no-good combinations that can never be part of a
model. In order to be able to do this, we need a few semantic predicates to
record the state of the calculation, as well as a system of disambiguating
automata, since it is possible to arrive at the same automaton by multiple
combinations of decisions and products. In \Catra{}, we use a number of
additional predicates to record the status of the materialisation of products,
to separate instances of our main predicate, and to register the mapping between
automata and their respective transition $\Filter$ terms, described in
\cref{sec:implementation}. However, at this point only rudimentary no-good
learning is implemented.

@@

Extending \Calculus{} to support fully symbolic automata is possible within the
framework, depending on your choice of $\Map$. The difficulty consists in
handling the mapping of the homomorphism $\Map$ over symbolic labels, assuming
it maps to finitely many monoid elements. This is not always straightforward, as
seen in the example of \cref{sec:multiple:example}. This complexity is inherent
in computing the full Parikh image, and stems from the fact that we need to
differentiate between the possible interpretations of the $\Sigma$ transitions
without knowing ahead of times which ones will actually be materialied in the
product. If, on the other hand, if our $\Map$ had been length-counting, which
does not differentiate between values, it would have required no adaptation at
all. With a somewhat liberal interpretation of what we are allowed to map to
(e.g. fresh terms), it is possible to use $\Map$ to represent choice operations
like the ones in a range label. In \Catra{}, we frontload this problem by
requiring the user to encode their input as a Parikh automaton. \Fudge{How is
the encoding of Ostrich Plus automata done?}

@@

Another interesting application is applying \Calculus{} to automata with
multiple tracks, e.g. transducers. One application of such a calculus would be
to represent replace operations and other functions on regular languages, and to
be able to answer questions like "does this operation change the length of the
string". The difficulty in implementing it for transducers is first to perform
the mapping on the labels, which for both the length and Parikh cases is
straightforward; just do the same thing in two dimensions. The more complicated
operation is defining the product of transducers. In some cases it would
probably be desirable to perform synchronisation (e.g. requiring overlapping
transitions) on only some transitions, for example the first track.

Implementing such a calculus is straightforward in \Calculus, since the
definition of products was left out of the definition. All that would be needed
is an appropriate update of the definitions of products. Simiarly, \Catra was
written with modularity in mind, and it should be straightforward to extend both
the input grammar and automata implementations to accomodate multi-track
automata with arbitrary synchronisation.

@@

The benchmarks are executed in parallel using GNU Parallel~\cite{parallel},
since they are mostly single-threaded and initial results showed negligible
interference on performance, on \BenchmarkRig{}. We compiled the code using
Scala~\ScalaVersion{}, and executed the experiments on~\JvmVersion{} with a
maximum heap of~\MaxHeapSize{}. We used \Nuxmv{} version~\NuxmvVersion{} invoked
as a subprocess for each instance. Experiments were executed on a fresh JVM per
backend, but the JVM was kept active for each batch of \BatchSize{} instances,
meaning that JIT compilation would make subsequent executions faster, while
still periodically restarting the JVM. We believe this represents a realistic
use case where \Calculus{} is integrated into a string solver or other system
and is called repeatedly. To mitigate systemic impacts of this decision on our
results, experiments were executed in random order for all backends. We run all
benchmarks with a timeout of~\RuntimeTimeout{}.

As a way of gauging the differences in overhead between the native backends and
\Nuxmv, we executed a small experiment where we ran the same trivial instance 30
times and observed the runtimes for all solvers. All backends saw an improvement
from JIT compilation within the first three runs, but subsequent runs only
improved the native backends. The final measured overhead for the fully warm
JVMs was less than~\OverheadSeconds. From this we draw the conclusion that major
differences in runtime between the native backends and \Nuxmv{} is more likely
to stem from innate differences rather than unfair overhead resulting from our
particular configuration. Additionally, one of the advantages of \Calculus{} to
\Nuxmv{} is that the former can me integrated into an automated theorem prover
while the latter has to be called as a separate process. In this sense, our
benchmarking setup can be said to be realistic.

@@

In Figures~\ref{fig:solve-division} we show how many of the~\NrBenchmarks{} the
respective back-ends could solve and with which status. A full summary of their
outcomes is also available in Table~\ref{tab:solve-status}. We see here that
\Calculus{} generally outperforms \Nuxmv{} on unsatisfiable instances, while
being slightly worse at satisfiable instances. Both \Nuxmv{} and \Calculus{}
massively and strictly outperform the baseline approach on every kind of
instance, but most of all on satisfiable instances. In fact, the baseline
approach was significantly harder to benchmark due to its much higher demand for
RAM.
